### 2017-07-02 motivation.tex
edits{

While reasoning in terms of anomalies is cumbersome, and as the above
example shows often inadequate, reasoning about weak isolation in
terms of low-level traces~\cite{adyaphd,gotsmanconcur15} complicates
high-level reasoning.  A possible alternative would interleave weak
isolation implementation details within the program, yielding a
(more-or-less) conventional concurrent program that can be then
subject to classical concurrent verification methods.  Considering the
size and complexity of real-world transaction systems, this strategy
is unlikely to scale.

@@

This simple characterization of RR isolation allows us to verify the
consistency condtions associated with the \C{new\_order}
transaction. First, since the database doesn't change ($\Delta' =
\Delta$) (where $\Delta'$ represents the state after the transaction
commits), during the execution of the transaction's body, we can
reason about \C{new\_order} as if it executed in complete isolation
until its commit point, leading to a verification process similar to
what would been applied when reasoning about serializability.  When
\C{new\_order} finishes execution and becomes ready to commit, a
transition that transfers the transaction'slocal writes ($\delta$) to
the unchanged database state ($\Delta$).  However, we are required to
consider the interference from concurrent transitions at this point,
which might change the database state from $\Delta$ to $\Delta_c$. If
this interference includes the effects of a concurrent \C{new\_order}
transaction (with the same \C{d\_id}), then verification fails as
described previously (Fig.~\ref{fig:new_order_execs}). Fortunately,
sequential reasoning shows that this is impossible - RR prevents a
concurrent \C{new\_order} transaction that modifies the same
\C{District} record as the current transaction (concretely, since the
record is already present in the current transaction's local log, any
transition from $\Delta$ to $\Delta_c$ cannot change this record).
Applying such axiomatic reasoning on \C{new\_order} allows us to prove
that the TPC-C invariant holds when the transaction is executed under
PostgreSQL's RR isolation.  Our proof framework generalizes this style
of reasoning to various isolation levels on databases.

The second observation that informs our approach is one that pertains
to automation. Program verification, even when machine-aided, often
entails significant annotation burden in the form of intermediary
assertions and loop invariants required to prove a program correct.
This is certainly true for concurrent program logics, such as
Rely-Gurantee, which extend Hoare logic with additional artifacts and
where (stable) intermediary assertions and loop invariants remain a
major source of annotation burden.  However, a relational database is
a significantly simpler abstraction than shared memory. There are no
pointers, or linked data structures, or aliasing.  Although a database
essentially abstracts a mutable state, the state is mutated through a
well-defined fixed number of interfaces (SQL statements), each tagged
with a logical formula describing what records are accessed and
updated.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as following:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariant, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The following two sections,
however, focus on laying theoretical foundations for reasoning about
weak isolation.

>>>>

While reasoning in terms of anomalies is cumbersome, and as the above
example shows often inadequate, reasoning about weak isolation in
terms of low-level traces~\cite{adyaphd,gotsmanconcur15} on memory
read and write actions complicates high-level reasoning.  A possible
alternative would interleave weak isolation implementation details
within the program, yielding a (more-or-less) conventional concurrent
program that can be then subject to classical concurrent verification
methods.  Considering the size and complexity of real-world
transaction systems, this strategy is unlikely to scale.

@@

This simple characterization of RR isolation allows us to verify the
consistency conditions associated with the \C{new\_order}
transaction. First, since the database doesn't change ($\Delta' =
\Delta$) (where $\Delta'$ represents the state after the transaction
commits), during the execution of the transaction's body, we can
reason about \C{new\_order} as though it executed in complete isolation
until its commit point, leading to a verification process similar to
what would been applied when reasoning about serializability.  When
\C{new\_order} finishes execution and becomes ready to commit, a
transition that transfers the transaction's local writes ($\delta$) to
the unchanged database state ($\Delta$).  However, we are required to
consider the interference from concurrent transitions at this point,
which might change the database state from $\Delta$ to $\Delta_c$. If
this interference includes the effects of a concurrent \C{new\_order}
transaction (with the same \C{d\_id}), then verification fails as
described previously (Fig.~\ref{fig:new_order_execs}). Fortunately,
sequential reasoning shows that this is impossible - RR prevents a
concurrent \C{new\_order} transaction that modifies the same
\C{District} record as the current transaction (concretely, since the
record is already present in the current transaction's local log, any
transition from $\Delta$ to $\Delta_c$ cannot change this record).
Applying such axiomatic reasoning on \C{new\_order} allows us to prove
that the TPC-C invariant holds when the transaction is executed under
PostgreSQL's RR isolation.  Our proof framework generalizes this style
of reasoning to various isolation levels on databases.

The second observation that informs our approach is one that pertains
to automation. Program verification, even when machine-aided, often
entails significant annotation burden in the form of intermediary
assertions and loop invariants required to prove a program correct.
This is certainly true for concurrent program logics, such as
Rely-Gurantee, which extend Hoare logic with additional artifacts and
where (stable) intermediary assertions and loop invariants remain a
major source of annotation burden.  However, a relational database is
a significantly simpler abstraction than shared memory. There are no
pointers, linked data structures, or aliasing.  Although a database
essentially abstracts a mutable state, the state is mutated through a
well-defined fixed number of interfaces (SQL statements), each tagged
with a logical formula describing what records are accessed and
updated.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as following:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariants, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The following two sections,
however, first focus on laying the theoretical foundations for
reasoning about weak isolation.
